---
title: "Spread of a disease: Coveying exponential data"
author: "Valdrich Fernandes"
date: "5/28/2020"
output:     
    ioslides_presentation:
        widescreen: true
        smaller: true

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Introduction

Most of us can't make intuative sense of an exponential spread but diseases spread exponentially. 

A simple depiction of the number of cases of a disease, while easy to understand doesn't speak much about the severity of the situation. 

Multipliers such as R~0~ (R-naught) do a better job at describing the spread, but it fails at conveying the idea to general population.

There must be a better way of conveying the facts. Maybe Doubling time?

## Doubling time | Merits and demerits {.columns-2}

**Time for the number of cases to double**

Merits:

- More intuative as larger numbers; means that it takes longer to double. Slower spread

- Represents exponential spread and its severity; better than the number of cases

- Can be calculated on a daily bases (after smoothing, like average of previous week)

- Better at predicting

<br>
<br>
Demerits:

- Slower to react to decrease in spread than the number of cases

- Prevelance of the disease is not represented

- Not widely adopted; might not be understandable at first glance

<br>

## Plot / Map {.columns-2}
```{r, fig.cap= 'Doubling time of COVID-19 in China and Italy. The vertical lines indicates 2 weeks since announcement of lockdown', out.height = '450px', out.width= '450px'}
library(dplyr, quietly = T, warn.conflicts = F)
library(stringr, quietly = T, warn.conflicts = F)
library(tidyr, quietly = T, warn.conflicts = F)
library(lubridate, quietly = T, warn.conflicts = F)
library(ggplot2, quietly = T, warn.conflicts = F)
library(ggpubr, quietly = T, warn.conflicts = F)

# download.file('https://opendata.ecdc.europa.eu/covid19/casedistribution/csv', 'Covid19.csv')
covid19 =  read.csv('Covid19.csv')
cases = pivot_wider(covid19, dateRep, names_from = countriesAndTerritories, values_from = cases)%>%
    mutate(dateRep = dmy(dateRep))
cases[is.na(cases)] = 0
cases = cases[order(cases$dateRep),]

ma = function(x, n = 5){
    stats::filter(x,rep(1 / n, n), sides = 1)
}

totCases = cbind(cases[,1],cumsum(cases[,-1]))
totCasesMA = apply(totCases[,-1], 2, ma)%>%
    as.data.frame%>%
    +1
totCasesMA = cbind(cases[,1],totCasesMA)

totCasesThresh = apply(totCasesMA[,-1], 2, function(x) x[which(x>50)])
totCasesThresh = totCasesThresh[lengths(totCasesThresh)>1]

CalcB = function(x, timelag = 1){
    1+diff(x,timelag)/x[(timelag+1):length(x)]
}

B = sapply(totCasesThresh, CalcB)
DoubleTime = sapply(B, function(b)log(2)/log10(b))
DoubleTime = DoubleTime[lengths(DoubleTime)>30]

gathering = function(.list, parameter = 'parameter'){
    Gathered = NULL
    for (i in c('Italy', 'China')){
        Gathered = rbind(Gathered, data.frame(days = 1:lengths(.list[i]),
                                              para = unlist(.list[i], use.names = F),
                                              Country = rep(i, times = lengths(.list[i]))))
    }
    names(Gathered)[2] = parameter
    Gathered
}

DoubleTimeGather = gathering(DoubleTime, 'DoubleTime')
LockdownIt = which(totCasesThresh$Italy < 4970&totCasesThresh$Italy > 4969)+14
LockdownCh = which(totCasesThresh$China < 540&totCasesThresh$China > 530)+14
p = ggplot(DoubleTimeGather, aes(days, DoubleTime, colour = Country))+
    geom_line(size = 1)+
    theme(legend.position = 'bottom')+ 
    labs(x = 'Days since 150 cases', y = 'Doubling time (days)')+
#    scale_y_log10()+
    coord_cartesian(ylim = c(1 ,800))+
    geom_vline(aes(xintercept = LockdownIt, colour = 'Italy'), linetype = 'dashed')+
    geom_vline(aes(xintercept = LockdownCh, colour = 'China'), linetype = 'dashed')+
    theme_bw()
totCasesGather = sapply(totCasesThresh, diff)%>%
    gathering('Cases')
p2 = ggplot(totCasesGather, aes(days, Cases, colour = Country))+
    geom_line(size = 1)+
    theme(legend.position = 'bottom')+ 
    labs(x = 'Days since 150 cases', y = 'Confirmed Cases')+
#    scale_y_log10()+
#    coord_cartesian(ylim = c(1 ,800))+
    geom_vline(aes(xintercept = LockdownIt, colour = 'Italy'), linetype = 'dashed')+
    geom_vline(aes(xintercept = LockdownCh, colour = 'China'), linetype = 'dashed')+
    theme_bw()
ggarrange(p, p2, nrow = 2, common.legend = TRUE)
```

- At the start; Few cases but short doubling time

- After Lockdown; Drop in number of cases but doubling time has just started increasing

## Summary {.bigger}

<div class = "centered">
**Early action is more effective than a late reaction**
</div>

- Doubling time can convey the severity of an exponential spread

- It can help inform government decisions during few cases

- It does a better job at informing citizens of the severity

- Can help identify countries at a higher risk